---
title: "Models"
author: "Minu Pabbathi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Libraries

```{r}
library(tidyverse)
library(tidymodels)
library(skimr) # skim()
library(corrr)
library(corrplot)
library(naniar) # vis_miss()
library(ROSE) #oversampling
library(discrim) #LDA
library(kernlab) # SVM kernels
tidymodels_prefer()
earthquakes <- read_csv("./data/earthquakes.csv")
earthquakes <- earthquakes %>% 
  mutate(alert = as.factor(alert), tsunami = as.factor(tsunami))
```

# Pre-processing

```{r}
earthquakes <- unique(earthquakes)
earthquakes <- select(earthquakes, -"postcode", -"url", -"detailUrl", -"what3words", -"locationDetails", -"type", -"ids", -"id", -"code", -"geometryType", -"place", -"title", -"placeOnly", -"timezone", -"sources", -"date", -"time", -"updated", -"status", -"magType", -"net", -"types", -"locality", -"city", -"country", -"subnational", -"continent", -"location")
```

# Splitting

```{r}
set.seed(42)
earthquakes_split <- initial_split(earthquakes, strata = tsunami)
earthquakes_train <- training(earthquakes_split)
earthquakes_test <- testing(earthquakes_split)
earthquakes_folds <- vfold_cv(earthquakes_train, v = 5, strata = tsunami)
```

# Recipe

```{r}
earthquakes_recipe <- recipe(tsunami ~ ., data = earthquakes_train) %>% 
  step_impute_mode(alert) %>%
  step_dummy(all_nominal_predictors()) %>% 
  step_center(all_predictors()) %>% 
  step_scale(all_predictors())
```

## Model Building

**explain models and the process of building the model**

### Logistic Model

```{r}
log_spec <- logistic_reg() %>% 
  set_mode("classification") %>% 
  set_engine("glm")

log_wf <- workflow() %>% 
  add_model(log_spec) %>% 
  add_recipe(earthquakes_recipe)

log_fit <- fit(log_wf, data = earthquakes_train)
```

ROC and AUC

```{r}
log_preds <- augment(log_fit, earthquakes_train)
log_preds %>%
  roc_curve(tsunami, .pred_1, event_level = "second") %>%
  autoplot()
```

```{r}
roc_auc(log_preds, truth = tsunami, .pred_1, event_level = "second")$.estimate
```

### LDA

```{r}
lda_spec <- discrim_linear() %>% 
  set_mode("classification") %>% 
  set_engine("MASS")

lda_wf <- workflow() %>% 
  add_model(lda_spec) %>% 
  add_recipe(earthquakes_recipe)

lda_fit <- fit(lda_wf, data = earthquakes_train)
```

ROC and AUC

```{r}
lda_preds <- augment(lda_fit, earthquakes_train)
lda_preds %>% 
  roc_curve(truth = tsunami, .pred_1, event_level = "second") %>% 
  autoplot()
```

```{r}
roc_auc(lda_preds, tsunami, .pred_1, event_level = "second")$.estimate
```

### KNN

```{r}
knn_spec <- nearest_neighbor(neighbors = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("kknn")

knn_wf <- workflow() %>% 
  add_model(knn_spec) %>% 
  add_recipe(earthquakes_recipe)

knn_grid <- grid_regular(neighbors(range = c(1, 10)),
                         levels = 10)
tune_knn <- tune_grid(knn_wf,
                      resamples = earthquakes_folds,
                      grid = knn_grid,
                      metrics = metric_set(roc_auc))
autoplot(tune_knn)

best_knn <- select_best(tune_knn, metric = "roc_auc")
knn_final_wf <- finalize_workflow(knn_wf, best_knn)
knn_final_fit <- fit(knn_final_wf, data = earthquakes_train)
```

ROC and AUC

this looks like a really good AUC but is slightly misleading since KNN is very prone to overfitting

```{r}
knn_preds <- augment(knn_final_fit, earthquakes_train)
knn_preds %>% 
  roc_curve(tsunami, .pred_1, event_level = "second") %>% 
  autoplot()
roc_auc(knn_preds, tsunami, .pred_1, event_level = "second")$.estimate
```

### Random Forest

```{r}
rf_spec <- rand_forest(mtry = tune(),
                       trees = tune(),
                       min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>% 
  set_mode("classification")

rf_wf <- workflow() %>% 
  add_model(rf_spec) %>% 
  add_recipe(earthquakes_recipe)

rf_grid <- grid_regular(mtry(range = c(1, 14)),
                        trees(range = c(100, 500)),
                        min_n(range = c(10, 20)),
                        levels = 14)
```

```{r, eval = F}
tune_rf <- tune_grid(rf_wf,
                     resamples = earthquakes_folds,
                     grid = rf_grid,
                     metrics = metric_set(roc_auc))
save(tune_rf, file = "tune_rf.rda")
```

Since this model takes a while to run, we can save it after running and load the results.

```{r}
load("tune_rf.rda")
autoplot(tune_rf)
```

```{r}
best_rf <- select_best(tune_rf, metric = "roc_auc")
rf_final_wf <- finalize_workflow(rf_wf, best_rf)
rf_final_fit <- fit(rf_final_wf, earthquakes_train)

rf_preds <- augment(rf_final_fit, earthquakes_train)

roc_curve(rf_preds, truth = tsunami, .pred_1, event_level = "second") %>% 
  autoplot()

conf_mat(rf_preds, truth = tsunami, .pred_class) %>% 
  autoplot(type = "heatmap")

roc_auc(rf_preds, truth = tsunami, .pred_1, event_level = "second")$.estimate
```

### Boosted Tree

```{r}
bt_spec <- boost_tree(trees = tune()) %>% 
  set_engine("xgboost") %>% 
  set_mode("classification")

bt_wf <- workflow() %>% 
  add_model(bt_spec) %>% 
  add_recipe(earthquakes_recipe)

bt_grid <- grid_regular(trees(range = c(10, 2000)),
                        levels = 10)

tune_bt <- tune_grid(bt_wf,
                     resamples = earthquakes_folds,
                     grid = bt_grid,
                     metrics = metric_set(roc_auc))

autoplot(tune_bt)

best_bt <- select_best(tune_bt, metric = "roc_auc")
bt_final_wf <- finalize_workflow(bt_wf, best_bt)
bt_final_fit <- fit(bt_final_wf, earthquakes_train)
```

ROC and AUC

```{r}
bt_preds <- augment(bt_final_fit, earthquakes_train)
roc_curve(bt_preds, tsunami, .pred_1, event_level = "second") %>% 
  autoplot()
roc_auc(bt_preds, tsunami, .pred_1, event_level = "second")$.estimate

conf_mat(bt_preds, truth = tsunami, .pred_class) %>% 
  autoplot(type = "heatmap")
```

### SVM Radial

```{r}
svm_rbf_spec <- svm_rbf(cost = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("kernlab")

svm_rbf_wf <- workflow() %>% 
  add_recipe(earthquakes_recipe) %>% 
  add_model(svm_rbf_spec)

svm_rbf_grid <- grid_regular(cost(), levels = 5)

tune_svm_rbf <- tune_grid(svm_rbf_wf,
                          resamples = earthquakes_folds,
                          grid = svm_rbf_grid,
                          metrics = metric_set(roc_auc))

 autoplot(tune_svm_rbf)

best_svm_rbf <- select_best(tune_svm_rbf, metric = "roc_auc")

svm_rbf_final_wf <- finalize_workflow(svm_rbf_wf, best_svm_rbf)
svm_rbf_final_fit <- fit(svm_rbf_final_wf, earthquakes_train)
```

ROC and AUC

```{r}
svm_rbf_preds <- augment(svm_rbf_final_fit, earthquakes_train)
roc_curve(svm_rbf_preds, tsunami, .pred_1, event_level = "second") %>% 
  autoplot()
roc_auc(svm_rbf_preds, tsunami, .pred_1, event_level = "second")$.estimate

conf_mat(svm_rbf_preds, truth = tsunami, .pred_class) %>% 
  autoplot(type = "heatmap")
```

### SVM Linear

```{r}
svm_lin_spec <- svm_poly(degree = 1, cost = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("kernlab")

svm_lin_wf <- workflow() %>% 
  add_recipe(earthquakes_recipe) %>% 
  add_model(svm_lin_spec)

svm_lin_grid <- grid_regular(cost(), levels = 5)

tune_svm_lin <- tune_grid(svm_lin_wf,
                          resamples = earthquakes_folds,
                          grid = svm_lin_grid,
                          metrics = metric_set(roc_auc))

 autoplot(tune_svm_lin)

best_svm_lin <- select_best(tune_svm_lin, metric = "roc_auc")

svm_lin_final_wf <- finalize_workflow(svm_lin_wf, best_svm_lin)
svm_lin_final_fit <- fit(svm_lin_final_wf, earthquakes_train)
```

ROC and AUC

```{r}
svm_lin_preds <- augment(svm_lin_final_fit, earthquakes_train)
roc_curve(svm_lin_preds, tsunami, .pred_1, event_level = "second") %>% 
  autoplot()
roc_auc(svm_lin_preds, tsunami, .pred_1, event_level = "second")$.estimate

conf_mat(svm_lin_preds, truth = tsunami, .pred_class) %>% 
  autoplot(type = "heatmap")
```
